Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-07 survey/survey.tex
Add \sumtype and \prodtype and make all Pi types use these macros. Fixed Upto Sec 3.

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inleft}[1]{\textsf{left}~#1}
\newcommand{\inright}[1]{\textsf{right}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

@@

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& t & \iso & t &: \idc \\
\\
\identlp :&  0 + t & \iso & t &: \identrp \\
\swapp :&  t_1 + t_2 & \iso & t_2 + t_1 &: \swapp \\
\assoclp :&  t_1 + (t_2 + t_3) & \iso & (t_1 + t_2) + t_3 &: \assocrp \\
\\
\identlt :&  1 * t & \iso & t &: \identrt \\
\swapt :&  t_1 * t_2 & \iso & t_2 * t_1 &: \swapt \\
\assoclt :&  t_1 * (t_2 * t_3) & \iso & (t_1 * t_2) * t_3 &: \assocrt \\
\\
\absorbr :&~ 0 * t & \iso & 0 ~ &: \factorzl \\
\dist :&~ (t_1 + t_2) * t_3 & \iso & (t_1 * t_3) + (t_2 * t_3)~ &: \factor
\end{array}
\]
\caption{$\Pi$-terms.}
\label{pi-terms}
\end{figure}

We now have in our hands our desired denotational semantics for types.
We want to create a programming language, which we call $\Pi$, such
that the types and type combinators map to
$\bot, \top, \uplus, \times$, and such that we have ground terms whose
denotation are all $16$ type isomorphisms of
Fig.~\ref{type-isos}. This is rather straightforward, as we can simply
do this literally. To make the analogy with commutative semirings
stand out even more, we will use $0, 1, +$, and $*$ at the type level,
and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
shows the ``constants'' of the language. Note how we have continued
with the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof
  terms}, but rather than being witnesses of inhabitation, they are
witnesses of equivalences. Thus we get an unexpected programming
language design:

@@

\begin{figure}[t]
\[
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_2 \iso t_3}
{\jdg{}{}{c_1 \odot c_2 : t_1 \iso t_3}}
{}
\qquad
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
{\jdg{}{}{c_1 \oplus c_2 : t_1 + t_3 \iso t_2 + t_4}}
{}
\]
\[
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
{\jdg{}{}{c_1 \otimes c_2 : t_1 * t_3 \iso t_2 * t_4}}
{}
\]
\caption{$\Pi$-combinators.}
\label{pi-combinators}
\end{figure}

@@

@@

\begin{figure}[t]
\[
\begin{array}{rrcll}
\identlsp :&  t + 0 & \iso & t &: \identrsp \\
\identlst :&  t * 1 & \iso & t &: \identrst \\
\\
\absorbl :&~ t * 0 & \iso & 0 ~ &: \factorzr \\
\distl :&~ t_1 * (t_2 + t_3) & \iso & (t_1 * t_2) + (t_1 * t_3)~ &: \factorl
\end{array}
\]
\caption{Additional $\Pi$-terms.}
\label{more-pi}
\end{figure}

@@

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1+b} & = & \idc \oplus\ \sem{b} \\
 \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
 \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp\, b_1 + b_2 } \\
 \sem{0 * b_2} & = & \absorbr \\
 \sem{1 * b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1 * b_2) * b_3} & = & \assocrt \odot \sem{b_1 * (b_2 * b_3)} \\
 \sem{(b_1+b_2) * b_3} & = & \dist \odot \sem{b_1 * b_3+b_2 * b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

\item
The product type \ensuremath{b_1 * b_2} may be represented both as one wire
labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

@@

\item
Sum types may similarly be represented using using parallel wires with
a \ensuremath{+} operator between them.

@@

\noindent
When tracing the execution of \ensuremath{b_1+b_2} represented by one
wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  When tracing
the execution of two additive wires, a value can reside on only one of
the two wires.

@@

\item
When representing complex types like \ensuremath{(b_1 * b_2)+b_3} some visual
grouping of the wires may be done to aid readability. The exact type
however will always be clarified by the context of the diagram.

@@

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1 * (b_2 * b_3)} or
\ensuremath{(b_1 * b_2) * b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories as well as
that for monoidal categories.

@@

\noindent
When tracing the execution of \ensuremath{b_1+b_2} represented by one wire, one
can think of a value of the form \ensuremath{\mathit{left} ~v_1} or
\ensuremath{\mathit{right} ~v_2} as flowing on the wire, where
\ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
tracking the flow of particles on the wires, one can verify that the
expected types for commutativity are satisfied.

@@

Distributivity and factoring are interesting because they represent
interactions between sum and pair types. Distributivity should
essentially be thought of as a multiplexer that redirects the flow of
\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1+b_2}, as shown
below.

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\[\begin{array}{rcl}
c & : & b * (1+1) \leftrightarrow b + b \\
c & = & \swapt \odot \dist \odot (\identlt \oplus \identlt)
\end{array}\]

@@

\paragraph*{Booleans}
Let us start with encoding booleans. We use the type \ensuremath{1+1} to
represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
Boolean negation is straightforward to define:

@@

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool} * (\mathit{bool} *  \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

@@

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  *  (\mathit{not}  *  \mathit{not})}

@@

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be
\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
then we essentially have an `if' expression.

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  *  b_1 \leftrightarrow \mathit{bool}  *  b_2}

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) + (\idc \otimes\  c_2)) \odot \factor}

@@

The diagram above shows the input value of type \ensuremath{(1+1) *  b_2}
processed by the distribute operator \ensuremath{\dist}, which converts it into
a value of type \ensuremath{(1 *  b_2)+(1 *  b_2)}. In the
\ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
the value of type~\ensuremath{b_1}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.

@@

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} *  \mathit{bool}\leftrightarrow \mathit{bool} *  \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} *  (\mathit{bool} *  \mathit{bool})\leftrightarrow \mathit{bool} *  (\mathit{bool} *  \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.

>>>>

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\sumtype}{+}
\newcommand{\prodtype}{\times}
\newcommand{\inleft}[1]{\textsf{left}~#1}
\newcommand{\inright}[1]{\textsf{right}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_{\sumtype}}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_{\sumtype}}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_{\sumtype}}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_{\prodtype}}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_{\prodtype}}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_{\prodtype}}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

%% \roshan{Use something else other than $*$ here to talk about
%%   idempotence.}

@@

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& t & \iso & t &: \idc \\
\\
\identlp :&  0 \sumtype t & \iso & t &: \identrp \\
\swapp :&  t_1 \sumtype t_2 & \iso & t_2 \sumtype t_1 &: \swapp \\
\assoclp :&  t_1 \sumtype (t_2 \sumtype t_3) & \iso & (t_1 \sumtype t_2) \sumtype t_3 &: \assocrp \\
\\
\identlt :&  1 {\prodtype} t & \iso & t &: \identrt \\
\swapt :&  t_1 {\prodtype} t_2 & \iso & t_2 {\prodtype} t_1 &: \swapt \\
\assoclt :&  t_1 {\prodtype} (t_2 {\prodtype} t_3) & \iso & (t_1 {\prodtype} t_2) {\prodtype} t_3 &: \assocrt \\
\\
\absorbr :&~ 0 {\prodtype} t & \iso & 0 ~ &: \factorzl \\
\dist :&~ (t_1 \sumtype t_2) {\prodtype} t_3 & \iso & (t_1 {\prodtype} t_3) \sumtype (t_2 {\prodtype} t_3)~ &: \factor
\end{array}
\]
\caption{$\Pi$-terms.}
\label{pi-terms}
\end{figure}

We now have in our hands our desired denotational semantics for types.
We want to create a programming language, which we call $\Pi$, such
that the types and type combinators map to
$\bot, \top, \uplus, \times$, and such that we have ground terms whose
denotation are all $16$ type isomorphisms of
Fig.~\ref{type-isos}. This is rather straightforward, as we can simply
do this literally. To make the analogy with commutative semirings
stand out even more, we will use $0, 1, \sumtype$, and ${\prodtype}$ at the type level,
and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
shows the ``constants'' of the language. Note how we have continued
with the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof
  terms}, but rather than being witnesses of inhabitation, they are
witnesses of equivalences. Thus we get an unexpected programming
language design:

@@

\begin{figure}[t]
\[
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_2 \iso t_3}
{\jdg{}{}{c_1 \odot c_2 : t_1 \iso t_3}}
{}
\qquad
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
{\jdg{}{}{c_1 \oplus c_2 : t_1 \sumtype t_3 \iso t_2 \sumtype t_4}}
{}
\]
\[
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
{\jdg{}{}{c_1 \otimes c_2 : t_1 {\prodtype} t_3 \iso t_2 {\prodtype} t_4}}
{}
\]
\caption{$\Pi$-combinators.}
\label{pi-combinators}
\end{figure}

@@

%% \roshan{Mixing types and numberix $+$ is super confusing. Let's use a
%%   block font of some sort for the numeric cases. There are only a few
%%   of them.}

@@

\begin{figure}[t]
\[
\begin{array}{rrcll}
\identlsp :&  t \sumtype 0 & \iso & t &: \identrsp \\
\identlst :&  t {\prodtype} 1 & \iso & t &: \identrst \\
\\
\absorbl :&~ t {\prodtype} 0 & \iso & 0 ~ &: \factorzr \\
\distl :&~ t_1 {\prodtype} (t_2 \sumtype t_3) & \iso & (t_1 {\prodtype} t_2) \sumtype (t_1 {\prodtype} t_3)~ &: \factorl
\end{array}
\]
\caption{Additional $\Pi$-terms.}
\label{more-pi}
\end{figure}

@@

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1{\sumtype}b} & = & \idc \oplus\ \sem{b} \\
 \sem{(b_1{\sumtype}b_2){\sumtype}b_3} & = & \assocrp \odot \sem{b_1 {\sumtype} (b_2 {\sumtype} b_3)} \\
 \sem{b_1 {\sumtype} b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp\, b_1 {\sumtype} b_2 } \\
 \sem{0 {\prodtype} b_2} & = & \absorbr \\
 \sem{1 {\prodtype} b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1 {\prodtype} b_2) {\prodtype} b_3} & = & \assocrt \odot \sem{b_1 {\prodtype} (b_2 {\prodtype} b_3)} \\
 \sem{(b_1{\sumtype}b_2) {\prodtype} b_3} & = & \dist \odot \sem{b_1 {\prodtype} b_3{\sumtype}b_2 {\prodtype} b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

\item
The product type \ensuremath{b_1 {\prodtype} b_2} may be represented both as one wire
labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

@@

\item
Sum types may similarly be represented using using parallel wires with
a \ensuremath{{\sumtype}} operator between them.

@@

\noindent
When tracing the execution of \ensuremath{b_1{\sumtype}b_2} represented by one
wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  When tracing
the execution of two additive wires, a value can reside on only one of
the two wires.

@@

\item
When representing complex types like \ensuremath{(b_1 {\prodtype} b_2){\sumtype}b_3} some visual
grouping of the wires may be done to aid readability. The exact type
however will always be clarified by the context of the diagram.

@@

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1 {\prodtype} (b_2 {\prodtype} b_3)} or
\ensuremath{(b_1 {\prodtype} b_2) {\prodtype} b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories as well as
that for monoidal categories.

@@

\noindent
When tracing the execution of \ensuremath{b_1{\sumtype}b_2} represented by one wire, one
can think of a value of the form \ensuremath{\mathit{left} ~v_1} or
\ensuremath{\mathit{right} ~v_2} as flowing on the wire, where
\ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
tracking the flow of particles on the wires, one can verify that the
expected types for commutativity are satisfied.

@@

Distributivity and factoring are interesting because they represent
interactions between sum and pair types. Distributivity should
essentially be thought of as a multiplexer that redirects the flow of
\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1{\sumtype}b_2}, as shown
below.

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\[\begin{array}{rcl}
c & : & b {\prodtype} (1{\sumtype}1) \leftrightarrow b {\sumtype} b \\
c & = & \swapt \odot \dist \odot (\identlt \oplus \identlt)
\end{array}\]

@@

\paragraph*{Booleans}
Let us start with encoding booleans. We use the type \ensuremath{1{\sumtype}1} to
represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
Boolean negation is straightforward to define:

@@

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool} {\prodtype} (\mathit{bool} {\prodtype}  \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

@@

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  {\prodtype}  (\mathit{not}  {\prodtype}  \mathit{not})}

@@

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1{\sumtype}b_2}. If we choose \ensuremath{1{\sumtype}1} to be
\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
then we essentially have an `if' expression.

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  {\prodtype}  b_1 \leftrightarrow \mathit{bool}  {\prodtype}  b_2}

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) {\sumtype} (\idc \otimes\  c_2)) \odot \factor}

@@

The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_2}
processed by the distribute operator \ensuremath{\dist}, which converts it into
a value of type \ensuremath{(1 {\prodtype}  b_2){\sumtype}(1 {\prodtype}  b_2)}. In the
\ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
the value of type~\ensuremath{b_1}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.

@@

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} {\prodtype}  \mathit{bool}\leftrightarrow \mathit{bool} {\prodtype}  \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})\leftrightarrow \mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.
